---
title: "STAT/MATH 495: Problem Set 09"
author: "Pei Gong"
date: "2017-11-07"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    collapsed: false
    smooth_scroll: false
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.width=8, fig.height=4.5, message=FALSE, warning = FALSE
  )
set.seed(76)

library(tidyverse)
```



# Collaboration

Please indicate who you collaborated with on this assignment: 



# Question 1: Run k-means

```{r}
observations_1 <- read_csv("data/observations_1.csv")
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_2

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

**Questions**:

1. Run KMC 10 times on `observations_1` and comment on the consistency of the
results.



1. Speculate on the root cause of any consistency or inconsistency in the
results.
1. Run KMC 10 times on `observations_2` and comment on the consistentcy of the results.
1. Speculate on the root cause of any consistency or inconsistency in the
results.

**Answers**:

1.
1.
1.
1.
```{r}
# Set observations to be one of two datasets
observations1 <- observations_1
clusters<-rnorm(10,1,1)
k_means_results<-rnorm(10,1,1)
cluster_centers<-rnorm(10,1,1)
k <- 2
k_means_results <- kmeans(observations1, centers=k)
  clusters <- k_means_results$cluster
  cluster_centers <- k_means_results$centers
  cluster_centers<- cluster_centers%>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

 # Run the additional 9 times through a for loop 

  for(i in 1:9){
  k_means_loop <- kmeans(observations1, centers=k)
  cluster_loop <- k_means_loop$centers
  cluster_centers_loop<- cluster_loop%>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))
  #combine new cluster centers to the exisiting cluster center 
  cluster_centersnew<-bind_rows(cluster_centers,cluster_centers_loop)
  }

```

Demonstrate the variation of the centers through standard deviation. 
```{r}
center1<-subset(cluster_centersnew,cluster==1)
center2<-subset(cluster_centersnew,cluster==2)

#for center 1
h_spread_center1<-sd(center1$x1);h_spread_center1
v_spread_center1<-sd(center1$x2);v_spread_center1  

#for center2
h_spread_center2<-sd(center2$x1);h_spread_center2
v_spread_center2<-sd(center2$x2);v_spread_center2  
```

Presenting the centers with the observations uncategorized. There are only 4 distinct outcomes from the 10 repeated clustering. 
```{r}
ggplot(NULL, aes(x=x1, y=x2)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centersnew, aes(col=cluster),size=5)

```

Now 10 times for observation 2 
```{r}
observations2 <- observations_2
k <- 2
k_means_results2 <- kmeans(observations2, centers=k)
  clusters2 <- k_means_results2$cluster
  cluster_centers2 <- k_means_results2$centers
  cluster_centers2<- cluster_centers2%>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

 # Run the additional 9 times through a for loop 
for(i in 1:9){
  k_means_loop2 <- kmeans(observations2, centers=k)
  cluster_loop2 <- k_means_loop2$centers
  cluster_centers_loop2<- cluster_loop2%>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))
  #combine new cluster centers to the exisiting cluster center 
  cluster_centersnew2<-bind_rows(cluster_centers2,cluster_centers_loop)
  }

```

Demonstrate the variation of the centers through standard deviation. 
```{r}
center1_2<-subset(cluster_centersnew2,cluster==1)
center2_2<-subset(cluster_centersnew2,cluster==2)

#for center 1
h_spread_center1_1<-sd(center1_2$x1);h_spread_center1_1
v_spread_center1_2<-sd(center1_2$x2);v_spread_center1_2 

#for center2
h_spread_center2_1<-sd(center2_2$x1);h_spread_center2
v_spread_center2_2<-sd(center2_2$x2);v_spread_center2  
```

Presenting the centers with the observations uncategorized. There are only 4 distinct outcomes from the 10 repeated clustering. 
```{r}
ggplot(NULL, aes(x=x1, y=x2)) +
  geom_point(data=observations2) +
  geom_point(data=cluster_centersnew2,aes(col=cluster),size=5)

```


# Bonus question: Code your own

Read ISLR page 388 Algorithm 10.1 and implement k-means clustering from scratch.
Don't worry about doing it for general $k$; keep it simple and do it for $k=2$
specifically. Apply it to `observations_2` from above.

```{r}
# Hint:
library(proxy)
A <- data_frame(
  x1 = c(0, 0.5, 0.75, 1),
  x2 = c(0, 0.5, 0.75, 1)
)
B <- data_frame(
  x1 = c(1, 0),
  x2 = c(1, 0)
)
distance_matrix <- proxy::dist(x=A, y=B)
distance_matrix
apply(distance_matrix, 1, which.min)
```
