---
title: "STAT/MATH 495: Problem Set 09"
author: "Pei Gong"
date: "2017-11-07"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    collapsed: false
    smooth_scroll: false
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.width=8, fig.height=4.5, message=FALSE, warning = FALSE
  )
set.seed(76)

library(tidyverse)
```

# Collaboration

Please indicate who you collaborated with on this assignment: Independent

# Question 1: Run k-means

```{r}
observations_1 <- read_csv("data/observations_1.csv")
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_2

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

**Questions**:

1. Run KMC 10 times on `observations_1` and comment on the consistency of the
results.
1. Speculate on the root cause of any consistency or inconsistency in the
results.
1. Run KMC 10 times on `observations_2` and comment on the consistentcy of the results.
1. Speculate on the root cause of any consistency or inconsistency in the
results.

**Answers**:

1.
```{r}
# Set observations to be one of two datasets
observations1 <- observations_1
k <- 2
k_means_results <- kmeans(observations1, centers=k)
  clusters <- k_means_results$cluster
  cluster_centers <- k_means_results$centers
  cluster_centers<- cluster_centers%>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

 # Run the additional 9 times through a for loop 
for(i in 1:9){
  k_means_loop <- kmeans(observations1, centers=k)
  cluster_loop <- k_means_loop$centers
  cluster_centers_loop<- cluster_loop%>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))
  #combine new cluster centers to the exisiting cluster center 
  cluster_centersnew<-bind_rows(cluster_centers,cluster_centers_loop)
  }

```

Presenting the centers with the observations uncategorized. There are only 1 pairs of distinct outcome from the 10 repeated clustering. The cluster center is pretty consistent. It might be that because of the distribution of the data, there is only one way of optimally assigning points to different clusters.  
```{r}
ggplot(NULL, aes(x=x1, y=x2)) +
  geom_point(data=observations1) +
  geom_point(data=cluster_centersnew, aes(col=cluster),size=5)

```

Now 10 times for observation 2 
```{r}
observations2 <- observations_2
k <- 2
k_means_results2 <- kmeans(observations2, centers=k)
  clusters2 <- k_means_results2$cluster
  cluster_centers2 <- k_means_results2$centers
  cluster_centers2<- cluster_centers2%>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

 # Run the additional 9 times through a for loop 
for(i in 1:9){
  k_means_loop2 <- kmeans(observations2, centers=k)
  cluster_loop2 <- k_means_loop2$centers
  cluster_centers_loop2<- cluster_loop2%>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))
  #combine new cluster centers to the exisiting cluster center 
  cluster_centersnew2<-bind_rows(cluster_centers2,cluster_centers_loop)
  }

```


Presenting the centers with the observations uncategorized. There are only 2 distinct pairs from the 10 repeated clustering. The cluster center is fairly consistent. Because of the distribution of the data, there are probably two ways of optimally assigning points to different clusters.  
```{r}
ggplot(NULL, aes(x=x1, y=x2)) +
  geom_point(data=observations2) +
  geom_point(data=cluster_centersnew2,aes(col=cluster),size=5)
```


# Bonus question: Code your own

Read ISLR page 388 Algorithm 10.1 and implement k-means clustering from scratch.
Don't worry about doing it for general $k$; keep it simple and do it for $k=2$
specifically. Apply it to `observations_2` from above.

```{r}
# Hint:
library(proxy)
A <- data_frame(
  x1 = c(0, 0.5, 0.75, 1),
  x2 = c(0, 0.5, 0.75, 1)
)
B <- data_frame(
  x1 = c(1, 0),
  x2 = c(1, 0)
)
distance_matrix <- proxy::dist(x=A, y=B)
distance_matrix
apply(distance_matrix, 1, which.min)
```
